A Study of the Effectiveness of Usage Examples in REST API Documentation

### 2017-03-22 controlled_study.tex
nearly all comments addressed from Craig

\begin{abstract}
Generating and maintaining REST API documentation with usage examples can be a time consuming and expensive process for evolving APIs. Most REST API documentation tools focus on automating the documentation of the API objects, but require manual effort for capturing realistic usage examples. Consequently, REST API developers need to know the cost vs. benefit of providing usage examples in the documentation as well as what to include in the usage examples to prioritize the documentation efforts. To this end, we have performed a controlled study with 26 experienced software engineers and found that REST API client developers face productivity problems with using correct data types, data formats, required HTTP headers and request body when documentation lacks usage examples. By following the REST API documentation guidelines from this paper, practitioners can reduce the errors, improve success rate and satisfaction of REST API client developers.

@@

\section{Introduction}
Researchers have emphasized API documentation as a key factor that impacts API usability both positively and negatively. To improve API documentation, researchers have recommended incorporating usage examples in the API documentation. REST APIs are used as the predominant application integration mechanism over the Internet. Generating and maintaining REST API documentation with usage examples can be an expensive process because most API documentation tools do not support effective usage examples.

In a resource constrained environment, it is important to understand the value of usage examples on REST API usability to allocate sufficient attention and efforts towards incorporating examples in the API documentation. While it is no surprise that examples help API client developers, REST API developers need to know what to include in the examples and why.

The documentation of REST APIs has distinctive features compared to the documentation of local APIs such as Java libraries. For example, local API documentation commonly comprises the description of classes and interfaces with their methods. In contrast, REST API documentation needs to include information about HTTP headers, request and response body and the data representation formats such as JSON, XML, etc. In our work we have focused on understanding the impact of usage examples within the realm of the distinctive REST API features.

We designed and performed a controlled study with experienced software engineers. Participants were divided into two groups and given the same set of API tasks to complete. One group was given the official WordPress REST API documentation and another group was given an enhanced version of the documentation where three usage examples were added. Using a novel technique, we collected 539 API calls made by the participants. We analyzed the data and observed recurring obstacles faced by the participants while performing the API tasks using the official documentation that lacks usage examples. Our contributions on this paper is as follows:

\begin{itemize}
  \item We provide a list of obstacles that REST API client developers face while performing API tasks using documentation that lacks usage examples.
  \item We provide a list of recommendations for REST developers to be used as a guideline to incorporate usage examples in the API documentation.
  \item We provide empirical evidence that usage examples in REST API documentation help API client developers perform API tasks with higher developer satisfaction, less time, and better success rate.
\end{itemize}

The remainder of this paper is organized as follows: In Section \ref{sec:method} we discuss our research goals, requirements, the object of the study and the participant selection, the study setup, data collection and analysis methods. In Section \ref{sec:results} we discuss the results of our analysis. We discuss the threats to validity and provide a summary of the related work in Sections \ref{sec:threat} and  \ref{sec:related_work} respectively.

\section{Methodology}
\label{sec:method}
\subsection{Research Goals}
This research is aimed at answering the following research questions:

@@

\subsection{Requirements} % (fold)

@@

\begin{itemize}
  \item \textbf{R1. Representative API.} We had to choose an existing REST API that is currently used by API client developers. Selecting a mature REST API for this study reduces the possibility that the obstacles we observe in the study are the results of insignificant accidental problems symptomatic of an immature technology. A familiar domain needed to be selected so that participants are able to relate to the API features without requiring upfront training. In addition to selecting the API, we had to select tasks that are related to the core features provided by the API to represent reality.
  \item \textbf{R2. Open source.} To be able to understand the impact of usage examples, we needed to select an open-source API where we can add new examples to the documentation for performing this study.
  \item \textbf{R3. Time bound.} The total time spent by each participant is required to be time bound to measure productivity. As a result, the study needs to be setup such that participants are able to focus on performing the tasks minimizing any overhead.
  \item \textbf{R4. Experienced developers.} Developers with prior experience on REST APIs need to be recruited as study participants to perform the study within a limited amount of time and in a realistic setup. Furthermore, to reduce a learning bias, only participants with no prior experince of using the WordPress REST API V2 are accepted for this study.
\end{itemize}

@@

We selected the WordPress REST API V2 for this study. WordPress is a blog-like open-source (R2) framework used by over 409 million people to visit 23.6 billion pages each month \footnote{\url{https://wordpress.com/activity/}}. The API allows programmatic access to list, create, update, and delete WordPress data such as blog posts, comments, users, images, tags, etc.

@@

Starting January 2017, the WordPress REST API V2 is no longer required to be installed as a separate plug-in since it's bundled with every WordPress installation. According to the code repository on GitHub, there are a total of 99 and 46 contributors that had at least one commit to the code repository of the API and its documentation, respectively. These properties satisfy R1, our requirements for using a representative API. By selecting an open-source project we are able to access the source-code to inspect the implementation and documentation technique of the WordPress REST API. The API implements a self-documenting feature where API developers expose API endpoints over HTTP OPTIONS verb to explain the API elements. To implement this feature, the API developers describe the API elements in the code. For example:

@@

On Listing \ref{wordpress_code}, line 4 specifies that this is a schema definition for the API element $Post$. On Line 10, it defines $date$, one of the properties of $Post$, followed by a human readable description and and type information. On line 14, the context of this property is mentioned as $view, edit, embed$, meaning that this property will be returned when the $Post$ object is returned, embedded, or can be used as an input for editing.

The WordPress team leverages the this self-documenting feature to generate and publish the official API documentation as an HTML website. Figure \ref{fig:schema} shows a screenshot of the published documentation for the $date$ attribute of the $Post$ API element.\footnote{\url{http://v2.wp-api.org/reference/posts/}}

@@

The participants are requested to perform a total of six tasks using the API including one practice task. All of the tasks are related to a single API element, $Post$ \footnote{A Post identifies a blog post within WordPress}. The tasks get progressively more difficult, and all but the last task can be performed independently of each other. Participants are requested to limit the total time on the study to a maximum of one hour. Participants are encouraged to proceed to the next task when they are either satisfied with their answer or feel stuck and unable to make progress.

@@

To verify the response of an API call, the API client developers can use HTTP response headers and/or HTTP response body. To perform the tasks, the participants are required to use one or more of the inputs I1-4. In the following paragraphs, we describe each task with its description and the study observation goals against the aforementioned API input and output information. For each of the tasks, the participants are required to use the same WordPress REST API and its documentation.

\textbf{T1: List all posts task}. We ask the participants to use the WordPress REST API to get a list of the blog posts from a given WordPress site. This is the practice task, and the inputs to answer this task are pre-filled for the participants. It allows the participants to understand the tools used for this study as well as get familiarity with the Post API. The answer for this task makes use of I1 and I2.

@@

\textbf{T3: Exclude posts by ids task.} We ask the participants to use the API to get a list of all posts excluding posts with IDs 1 and 4. Participants need to use the inputs I1-2, and use a desired format on I2 to pass an array of IDs as a parameter. This task allows us to understand how API client developers identify the format for using an array within the URL with respect to the usage examples in the API documentation.

@@

A total 26 participants from sixteen different companies and six different countries (Canada, USA, Germany, Ireland, Brazil, and,
Bangladesh) were recruited through online announcements posted on Twitter, Facebook, and software developer focused mailing lists. Table \ref{table:participants} shows a summary of the experience level of the participants in each group.

@@

We conducted two pilot studies to evaluate and understand an effective process for performing this study. The first pilot study involved four participants that were invited to join the first author in-person or over video conferencing. The study involved tasks using two APIs, the WordPress REST API V2 and the GMail REST API. Each participant was given one of the two APIs and a set of tasks and online answer forms to complete using the API within an hour. The primary findings from this pilot are as follows: 1) participants required significant overhead time to setup a development environment with the proper API credentials, 2) the intermediate trial attempts of using the API are potentially more valuable than the final answer as it allows us to study what obstacles the participants face, 3) the number of tasks for the study had to be reduced to fit within the one hour limit, and 4) for GMail API, participants used up a large portion of their time on learning about how to use  OAuth.

\begin{figure*}[t]
\begin{center}
\fbox{\includegraphics[width=\linewidth,keepaspectratio]{api_explorer.png}}
\end{center}
\caption{Screenshot of the web-based REST API explorer}
\label{fig:api_explorer}
\end{figure*}

To overcome the shortcomings found from the first pilot study, we decided to develop a web-based REST API explorer as shown on Figure \ref{fig:api_explorer} that allows participants to use their browser to make the API calls without setting up any development environment. The web-based API explorer only requires the inputs I1-4, and displays the HTTP response headers and body on the click of a button. Thus the participants could focus on using the right input and verifying the output without having to write any code. The web-base REST API explorer also allowed us to automatically collect all the trial API calls that the participants make for each API task. A second pilot study involving seven new participants was performed to collect usability related feedback about the web-based REST API explorer. We only used WordPress since it didn't require knowledge about OAuth. From this study, we observed patterns of mistakes that API client developers make that can be reduced by adding usage examples in the API documentation.

\subsubsection{Protocol}

Learning from the pilot studies, we designed the actual study protocol as follows: we enhanced the original WordPress REST API documentation and added a total of 3 API usage examples to show listing of blog posts with query parameters for filtering, a request to create a blog post and a request to update a blog post. We used the WordPress API unit tests to find relevant data for these examples. Figure \ref{fig:original} shows a screenshot of the original API documentation related to T4 where the API client developers are provided with a reference table describing the different properties that can be used to create a $Post$ object\footnote{\url{http://v2.wp-api.org/reference/posts/}}. Figure \ref{fig:enhanced} shows a screenshot of the enhanced API documentation with a cURL based usage example. cURL is used because it's used by elsewhere in the original API documentation. In the enhanced API documentation, the example shows one possible API call with realistic values for the data that is described in the reference table and associated API response headers and body.

We divided the participants into two groups, G1 and G2. G1 participants were provided with a link to the official API documentation on the web-based API explorer, and G2 participants were provided with a link to the enhanced API documentation with usage examples. The web-based API explorer allocated more participants to G1 compared to G2 because we wanted to better understand the impact of the lack of usage examples on API client developer productivity. However, each individual participant was randomly assigned to a group by the web-based API explorer. All participants were given the same set of API tasks and were requested to limit their participation time to a maximum of one hour. No task specific time limit was imposed except an overall limit one hour for the entire study because we wanted participants to spend sufficient time on each task without forcing them to move the next one. Participants were allowed to access the internet and external resources alongside the provided API documentation to perform the tasks as they would normally use on a typical work day. We used the web-based API explorer to also collect an experience rating on a scale of 0-10, 10 being the best possible, of using the given REST API documentation and a free-form feedback from each participant as an exit survey.

@@

For each API task attempted by each participant, we annotated the artifact with one of the following labels: successful, partially successful, and unsuccessful. Task participations are marked successful when I1-4 matches the required values for performing the given task. If a participant is able to use the correct I1-4 for one of the two tasks required to complete a single task (T2), we marked it as a partially successful. Otherwise, it's marked as unsuccessful. Based on these annotations, the following formulas were used to compute the quantitative results:

Success Rate = (total number of successful tasks by group) / (number of participants in group * number of API tasks)

Average Trial API Calls = (number of API calls annotated as successful) / (total number of trial API calls made by group)

Average Time Spent = (number of successful API tasks by group) / (total time spent by all participants in the group)

@@

We present a summary of the quantitative results as found by analyzing a total of 539 API calls (385 from G1, 152 from G2) made by the participants from this study in Figure \ref{fig:comparison}. Figure \ref{graph:productivity} juxtaposes the average number of trial API calls made by each group against their rate of success in performing the given tasks T2-T6. The average number of trial API calls and success rate for each group is computed using the following formulas:

Even though we expected G2 to outperform G1, we were surprised to see the difference in effort vs. success of the two groups. In Figure \ref{graph:productivity} and \ref{graph:time} we see that, for the simpler tasks, T2 and T3, G1 required more time and more trial API calls and reached successful answers with a lower percentage compared to G2. For T4, where the participants are required to inspect an HTTP response header, G2 needed more API calls on average, but also yielded 24\% higher success rate. For the more complex tasks T5 and T6, we see that participants in G1 had very low success rate and very high number of average API trials compared to G2. The data here depicts that without usage examples developers spend more time and require more trial and error attempts yet resulting in lower success rates, and the situation gets worse for relatively complicated tasks.

@@

In Figure \ref{fig:rating}, we see that the G2 participants provided higher satisfaction rating compared to G1 participants. P1.1 provided a satisfaction rating of 9, and analyzing the participant feedback against the success, we see a false positive case, where the participant perceived that the tasks were successfully completed even though only 2 of the 5 tasks were correctly answered. Other than this case, the satisfaction ratings provided by G1 participants consistently fall behind the G2 participants' ratings.

@@

\textbf{Observation: Use of Request Headers.} We were surprised to see that 14 of the 15 G1 participants that tried T5 and T6 faced problem using correct request headers without usage examples. Tasks T5 and T6 required the use of a HTTP request header named $Content-Type$. In the original API documentation, the following was mentioned: ``The API uses JSON exclusively as the request and response format, including error responses.''. Experienced software engineers still had problem using this HTTP header as we found from the participants in G1. P1.12 mentioned the following: \textit{``I didn't know how to specify application/json so that I could send POST data via the body instead in the URL parameters... Examples would really help - with a sample request and a sample response.''}  Even though G1 participants have years of experience with REST APIs, only one of them could use the $Content-Type$ header correctly without example. In contrast, 2 out of the 10 participants from G2 failed used this header correctly as indicated in the example in the enhanced documentation. Moreover, 1 of the 2 G2 participants that didn't use the header first time, later used it correctly on a subsequent attempt.

@@

\section{Threats to Validity}
\label{sec:threat}
  The selection of the API, the tasks, and the participants may introduce selection bias. We have developed and followed a set of requirements as discussed earlier on this paper to reduce this bias. This study needs to be replicated with different sets of APIs, tasks, and participants to generalize the results. The analysis of the raw data may introduce a bias. To reduce this, we have involved multiple researchers in the analysis process.

@@

\subsubsection{API Learning Obstacles}
Robillard et al. used a mixed approach of surveys and in-person interviews with professional software engineers to understand the API learning obstacles \cite{Robillard_what_makes} \cite{Robillard_a_field_study}. They found that many obstacles for API learnabilty were related to API documentation. They identified five impactful factors for API documentation as follows: documentation of intent, code examples, matching APIs with scenarios, penetrability of the API, and format and presentation. Uddin and Robillard surveyed 323 professional software engineers to understand the different ways in which a piece of documentation can be unfit for purpose \cite{g_uddin}. They identified ambiguity, incompleteness, and incorrectness as the three severest problems that lead API documentations to fail to answer the information needs of developers. Duala-Ekoko and Robillard performed a study with 20 programmers to understand the types of questions that developers have while using an unfamiliar API \cite{Duala-Ekoko:2012:AAQ:2337223.2337255}. They categorized the API related information needs into twenty generic questions that can be used to analyze the quality of API documentation. Myers et al. found that participants had limited success finding the relevant API elements to perform an API task using an enterprise API \cite{Myers_study}. Among other factors, they recommended providing code examples in the API documentation to help API client developers. Our work fits in the general space of studies of API documentation obstacles by focusing specifically on problems related to the absence of examples, and by expanding into the study of REST APIs.

\subsubsection{Crowd-Sourced API Usage Examples}
Wang et al. performed an exploratory study to understand the current state of API related knowledge available on the Internet \cite{6462686}. They searched the web for usage examples related to the API of five popular Java libraries and found that on an average API examples could be found for 77\% of the 4,637 APIs included in the selected libraries. Moreover, they found that the crowd-sourcing sites accounted for 93.7\% of the usage examples compared to 6.13\% that were published on the official documentation. An evaluation of Jadite showed that developers were three times faster to complete API tasks with access to auto-referred crowd-sourced usage examples~\cite{5295283}. Nasehi et al. analyzed StackOverflow threads to understand the characteristics of good and bad code examples based on user provided votes \cite{Nasehi_what_makes}. They found that commonly down-voted API related answers lacked code examples, explanations, and shortcoming of solutions. Treude et al. proposed a machine learning based approach to find relevant API documentation from StackOverflow using both the text and metadata found from the questions and answers~\cite{Treude:2016:AAD:2884781.2884800}. Kim et al. presented a technique to automatically augment code examples from the web using code search tools to Java APIs \cite{Kim:2009:AEJ:1747491.1747552}. Jiau et al. observed a severe inequality within the context of crowd-sourced API usage examples where most of the content were related to the popular APIs \cite{jiau}. Existing research on crowd-sourced API documentation has mostly focused finding usage examples of local APIs. In this paper, we focused on REST APIs, where the API client development programming language is agnostic to API implementation language. This distinction with local APIs makes it hard to leverage the techniques proposed in the aforementioned papers for finding crowd-sourced REST API usage examples.

\subsubsection{Measuring API Usability}
Rama et al. presented a set of formulas for computing a measure of API usability based on the API's structural components such as classes, methods, parameters, return values, thread-safety, etc~\cite{rama}. Scheller et al. presented a framework for automatically measuring the complexity of an API~\cite{scheller2015automated}. They identified a list of measurable API properties and provided formulas to compute complexities related to the interfaces, implementation and setup of an API. Grill et al. suggested using a combination of expert opinion and developer workshops to identify and collect feedback about API and its documentation related problems \cite{Grill2012}. Farooq et al. recommended using peer reviews of API code in addition to API usability studies to uncover API related bugs and incorporate feedback~\cite{Farooq:2010:AUP:1753326.1753677}. In our work, instead of measuring API usability, we have focused specifically on the relationship of REST API usability with respect to usage examples.

>>>>

\begin{abstract}
Generating and maintaining REST API documentation with usage examples can be a time consuming and expensive process for evolving APIs. Most REST API documentation tools focus on automating the documentation of the API objects, but require manual effort for capturing usage examples. Consequently, REST API developers need to know the cost vs. benefit of providing usage examples in the documentation to prioritize the documentation efforts. To this end, we have performed a controlled study with 26 experienced software engineers to understand problems that REST API client developers face while using an API without usage examples. We found that REST API client developers face productivity problems with using correct data types, data formats, required HTTP headers and request body when documentation lacks usage examples. By following the REST API documentation guidelines from this paper, REST API developers can reduce the errors, improve success rate and satisfaction of API client developers.

@@

\section{Introduction}
REST APIs are used as the predominant application integration mechanism over the Internet. Generating and maintaining REST API documentation with usage examples can be an expensive process because most API documentation tools do not support effective usage examples. Researchers have emphasized API documentation as a key factor that impacts API usability both positively and negatively. To improve API documentation, researchers have recommended incorporating usage examples in the API documentation.

In a resource constrained environment, it is important to understand the value of usage examples on REST API usability to allocate sufficient attention and efforts towards incorporating examples in the API documentation. While it is expected that examples help developers, REST API developers need to know what to include in the examples and why.

The documentation of REST APIs has distinctive features compared to the documentation of local APIs such as Java libraries. For example, local API documentation commonly comprises the description of classes and interfaces with their methods. In contrast, REST API documentation needs to include information about HTTP headers, request and response body and the data representation formats such as JSON, XML. The existing research on API usability areas have primarily focused on local APIs. In our work we have focused on understanding the impact of usage examples within the realm of the distinctive REST API features.

We designed and performed a controlled study with experienced software engineers to understand how REST API client developers are affected while using an API documentation that describes the API elements but lacks usage examples. Participants were divided into two groups and given the same set of API tasks to complete. One group was given the official WordPress REST API documentation and another group was given an enhanced version of the documentation where three usage examples were added. Using a novel technique, we collected 539 API calls made by the participants. We analyzed the data and observed recurring obstacles faced by the participants while performing the API tasks using the official documentation that lacks usage examples. Our contributions on this paper are as follows:

\begin{itemize}
  \item We provide a list of obstacles that REST API client developers face while performing API tasks using documentation that lacks usage examples.
  \item We provide a list of recommendations for REST developers to be used as a guideline to incorporate usage examples in API documentation.
  \item We provide empirical evidence that usage examples in REST API documentation help API client developers perform API tasks with higher developer satisfaction, less time, and better success rate.
\end{itemize}

The remainder of this paper is organized as follows: In Section \ref{sec:goal} we present our research question. In Section \ref{sec:method} we discuss our methodology in terms of the study requirements, the selected API of the study and the participant selection, the study setup, data collection and analysis methods. In Section \ref{sec:results} we discuss the results of our analysis. We discuss the threats to validity and provide a summary of the related work in Sections \ref{sec:threat} and  \ref{sec:related_work} respectively.

\section{Research Question}
\label{sec:goal}
This research is aimed at answering the following research question:

@@

\section{Methodology} % (fold)
\label{sec:method}

@@

\begin{itemize}
  \item \textbf{R1. Representative API.} We had to choose an existing REST API that is currently used by API client developers. Selecting a mature REST API for this study reduces the possibility that the obstacles we observe in the study are the results of insignificant accidental problems symptomatic of an immature technology. A familiar domain needed to be selected so that participants are able to relate to the API features without requiring upfront training. In addition to selecting the API, we had to select tasks that are related to the core features provided by the API to represent reality.
  \item \textbf{R2. Open source.} To be able to understand the impact of usage examples, we needed to select an open-source API where we can add new examples to the documentation for performing this study.
  \item \textbf{R3. Time bound.} The total time spent by each participant is required to be time bound to measure the rate of success within a limited time frame. As a result, the study needs to be setup such that participants are able to focus on performing the tasks minimizing any overhead.
  \item \textbf{R4. Participant Selection.} Developers with prior experience on REST APIs need to be recruited as study participants to perform the study within a limited amount of time and in a realistic setup. Furthermore, to reduce a learning bias, only participants with no prior experince of using the WordPress REST API V2 are accepted for this study.
\end{itemize}

@@

We selected the WordPress REST API V2 for this study. WordPress is a blog-like open-source (R2) framework used by over 409 million people to visit 23.6 billion pages each month \footnote{\url{https://wordpress.com/activity/}}. The API allows programmatic access to list, create, update, and delete WordPress data such as blog posts, comments, users, images, tags.

@@

Starting January 2017, the WordPress REST API V2 is no longer required to be installed as a separate plug-in since it is bundled with every WordPress installation. According to the code repository on GitHub, there are a total of 99 and 46 contributors that had at least one commit to the code repository of the API and its documentation, respectively. These properties satisfy R1, our requirements for using a representative API. By selecting an open-source project we are able to access the source-code to inspect the implementation and documentation technique of the WordPress REST API. The API implements a self-documenting feature where API developers expose API endpoints over HTTP OPTIONS verb to explain the API elements. To implement this feature, the API developers describe the API elements in the code. For example:

@@

On Listing \ref{wordpress_code}, line 4 specifies that this is a schema definition for the API element $Post$. On Line 10, it defines $date$, one of the properties of $Post$, followed by a human readable description and type information. On line 14, the context of this property is mentioned as $view, edit, embed$, meaning that this property will be returned when the $Post$ object is returned, embedded, or can be used as an input for editing.

The WordPress team leverages this self-documenting feature to generate and publish the official API documentation as an HTML website. Fig. \ref{fig:schema} shows a screenshot of the published documentation for the $date$ attribute of the $Post$ API element.\footnote{\url{http://v2.wp-api.org/reference/posts/}}

@@

The participants are requested to perform a total of six tasks using the API including one practice task. All of the tasks are related to a single API element, $Post$\footnote{A Post identifies a blog post within WordPress}. The tasks get progressively more difficult, and all but the last task can be performed independently of each other. Participants are requested to limit the total time on the study to a maximum of one hour. Participants are encouraged to proceed to the next task when they are either satisfied with their answer or feel stuck and unable to make progress.

@@

To verify the response of an API call, the API client developers can use HTTP response headers and/or HTTP response body. To perform the tasks, the participants are required to use one or more of the inputs I1-4. In the following paragraphs, we describe each task with its description and the study observation goals against the aforementioned API input and output information. For each of the tasks, the participants are required to use the same WordPress REST API and accompanying documentation.

\textbf{T1: List all posts task}. We asked the participants to use the WordPress REST API to get a list of the blog posts from a given WordPress site. This is the practice task, and the inputs to answer this task are pre-filled for the participants. It allows the participants to understand the tools used for this study as well as get familiarity with the Post API. The answer for this task makes use of I1 and I2.

@@

\textbf{T3: Exclude posts by IDs task.} We ask the participants to use the API to get a list of all posts excluding posts with IDs 1 and 4. Participants need to use the inputs I1-2, and use a desired format on I2 to pass an array of IDs as a parameter. This task allows us to understand how API client developers identify the format for using an array within the URL with respect to the usage examples in the API documentation.

@@

\begin{figure*}[t]
\begin{center}
\fbox{\includegraphics[width=\linewidth,keepaspectratio]{api_explorer.png}}
\end{center}
\caption{Screenshot of the Custom-built Web-based REST API Explorer Used by Study Participants}
\label{fig:api_explorer}
\end{figure*}

We recruited a total of 26 participants from sixteen different companies and six different countries (Canada, USA, Germany, Ireland, Brazil, and,
Bangladesh) through online announcements posted on Twitter, Facebook, and software developer focused mailing lists. Table \ref{table:participants} shows a summary of the experience level of the participants in each group.

@@

We conducted two pilot studies to evaluate and understand an effective process for performing this study. The first pilot study involved four participants that were invited to join the first author in-person or over video conferencing. The study involved tasks using two APIs, the WordPress REST API V2 and the GMail REST API. Each participant was given one of the two APIs and the tasks T1-T6 and online answer forms to complete using the API within an hour. The primary findings from this pilot are as follows: 1) participants required significant overhead time to setup a development environment with the proper API credentials, 2) the intermediate trial attempts of using the API are potentially more valuable than the final answer as it allows us to study what obstacles the participants face, 3) the number of tasks for the study had to be reduced to fit within the one hour limit, and 4) for the GMail API, participants used up a large portion of their time on learning about how to use  OAuth.

To overcome the shortcomings found from the first pilot study, we decided to develop a web-based REST API explorer as shown on Fig. \ref{fig:api_explorer} that allows participants to use their browser to make the API calls without setting up any development environment. The web-based API explorer only requires the inputs I1-4, and displays the HTTP response headers and body on the click of a button. Thus the participants could focus on using the right input and verifying the output without having to write any code. The web-base REST API explorer also allowed us to automatically collect all the trial API calls that the participants make for each API task. A second pilot study involving seven new participants was performed to collect usability related feedback about the web-based REST API explorer. We only used WordPress since it didn't require knowledge about OAuth. From this study, we observed patterns of mistakes that API client developers make that can be reduced by adding usage examples in the API documentation.

\subsubsection{Protocol}
\label{sub:protocol}
Learning from the pilot studies, we designed the main study protocol as follows: we enhanced the original WordPress REST API documentation and added a total of 3 API usage examples to show listing of blog posts with query parameters for filtering, a request to create a blog post and a request to update a blog post. We used the WordPress API unit tests to find relevant data for these examples. Fig. \ref{fig:original} shows a screenshot of the original API documentation related to T4 where the API client developers are provided with a reference table describing the different properties that can be used to create a $Post$ object\footnote{\url{http://v2.wp-api.org/reference/posts/}}. Fig. \ref{fig:enhanced} shows a screenshot of the enhanced API documentation with a cURL based usage example. cURL is used because it is used by elsewhere in the original API documentation. In the enhanced API documentation, the example shows one possible API call with realistic values for the data that is described in the reference table and associated API response headers and body.

We divided the participants into two groups, G1 and G2. G1 participants were provided with a link to the official API documentation on the web-based API explorer, and G2 participants were provided with a link to the enhanced API documentation with usage examples. The web-based API explorer allocated more participants to G1 compared to G2 because we wanted to better understand the impact of the lack of usage examples on API client developer productivity. However, each individual participant was randomly assigned to a group by the web-based API explorer. All participants were given the same set of API tasks and were requested to limit their participation time to a maximum of one hour. No task specific time limit was imposed except an overall limit one hour for the entire study because we wanted participants to spend sufficient time on each task without forcing them to move to the next one. Participants were allowed to access the internet and external resources alongside the provided API documentation to perform the tasks as they would normally use on a typical work day. We used the web-based API explorer to also collect an experience rating on a scale of 0-10, 10 being the best possible, of using the given REST API documentation and a free-form feedback from each participant as an exit survey.

@@

For each API task attempted by each participant, we annotated the artifact with one of the following labels: successful, partially successful, and unsuccessful. Task participations are marked successful when I1-4 matches the required values for performing the given task. If a participant is able to use the correct I1-4 for one of the two tasks required to complete a single task (T2), we marked it as partially successful. Otherwise, it's marked as unsuccessful. Based on these annotations, the following formulas were used to compute the quantitative results:

\textbf{Success Rate} = (total number of successful tasks by group) / (number of participants in group * number of API tasks)

\textbf{Average Trial API Calls} = (number of API calls annotated as successful) / (total number of trial API calls made by group)

\textbf{Average Time Spent} = (number of successful API tasks by group) / (total time spent by all participants in the group)

@@

We present a summary of the quantitative results as found by analyzing a total of 539 API calls (385 from G1, 152 from G2) made by the participants from this study in Fig. \ref{fig:comparison}. Fig. \ref{graph:productivity} juxtaposes the average number of trial API calls made by each group against their rate of success in performing the given tasks T2-T6. The average number of trial API calls and success rate for each group is computed using the following formulas:

Even though we expected G2 to outperform G1, we were surprised to see the difference in effort vs. success of the two groups. In Fig. \ref{graph:productivity} and \ref{graph:time} we see that, for the simpler tasks, T2 and T3, G1 required more time and more trial API calls and reached successful answers with a lower percentage compared to G2. For T4, where the participants are required to inspect an HTTP response header, G2 needed more API calls on average, but also yielded 24\% higher success rate. For the more complex tasks T5 and T6, we see that participants in G1 had very low success rate and very high number of average API trials compared to G2. The data here depicts that without usage examples developers spend more time and require more trial and error attempts yet resulting in lower success rates, and the situation gets worse for relatively complicated tasks.

@@

In Fig. \ref{fig:rating}, we see that the G2 participants provided higher satisfaction rating compared to G1 participants. P1.1 provided a satisfaction rating of 9, and analyzing the participant feedback against the success, we see a false positive case, where the participant perceived that the tasks were successfully completed even though only 2 of the 5 tasks were correctly answered. Other than this case, the satisfaction ratings provided by G1 participants consistently fall behind the G2 participants' ratings.

@@

\textbf{Observation: Use of Request Headers.} We were surprised to see that 14 of the 15 G1 participants that tried T5 and T6 faced problem using correct request headers without usage examples. Tasks T5 and T6 required the use of a HTTP request header named $Content-Type$. In the original API documentation, the following was mentioned: ``The API uses JSON exclusively as the request and response format, including error responses.''. Experienced software engineers still had problem using this HTTP header as we found from the participants in G1. P1.12 mentioned the following: \textit{``I didn't know how to specify application/json so that I could send POST data via the body instead in the URL parameters... Examples would really help - with a sample request and a sample response.''}  Even though G1 participants have years of experience with REST APIs, only one of them could use the $Content-Type$ header correctly without example. In contrast, 2 out of the 10 participants from G2 failed to use this header correctly as indicated in the example in the enhanced documentation. Moreover, 1 of the 2 G2 participants that didn't use the header first time, later used it correctly on a subsequent attempt.

@@

\section{Threats to Validity}
\label{sec:threat}
  The selection of the API, the tasks, and the participants may introduce selection bias. We have developed and followed a set of requirements as discussed earlier on this paper to reduce this bias (see Section \ref{sub:protocol}). This study needs to be replicated with different sets of APIs, tasks, and participants to generalize the results. The analysis of the raw data may introduce a bias. To reduce this, we have involved multiple researchers in the analysis process.

@@

\subsubsection{API Learning Obstacles}
Robillard et al. used a mixed approach of surveys and in-person interviews with professional software engineers to understand the API learning obstacles \cite{Robillard_what_makes} \cite{Robillard_a_field_study}. They found that many obstacles for API learnabilty were related to API documentation. They identified five impactful factors for API documentation as follows: documentation of intent, code examples, matching APIs with scenarios, penetrability of the API, and format and presentation. Uddin and Robillard surveyed 323 professional software engineers to understand the different ways in which a piece of documentation can be unfit for purpose \cite{g_uddin}. They identified ambiguity, incompleteness, and incorrectness as the three severest problems that lead API documentations to fail to answer the information needs of developers. Duala-Ekoko and Robillard performed a study with 20 programmers to understand the types of questions that developers have while using an unfamiliar API \cite{Duala-Ekoko:2012:AAQ:2337223.2337255}. They categorized the API related information needs into twenty generic questions that can be used to analyze the quality of API documentation. Myers et al. found that participants had limited success finding the relevant API elements to perform an API task using an enterprise API \cite{Myers_study}. Among other factors, they recommended providing code examples in the API documentation to help API client developers. Our work fits in the general space of studies of API documentation obstacles by focusing specifically on problems related to the absence of examples, and by expanding into the area of REST APIs.

\subsubsection{Crowd-Sourced API Usage Examples}
Wang et al. performed an exploratory study to understand the current state of API related knowledge available on the Internet \cite{6462686}. They searched the web for usage examples related to the API of five popular Java libraries and found that on average API examples could be found for 77\% of the 4,637 APIs included in the selected libraries. Moreover, they found that the crowd-sourcing sites accounted for 93.7\% of the usage examples compared to 6.13\% that were published on the official documentation. An evaluation of Jadite showed that developers were three times faster to complete API tasks with access to auto-referred crowd-sourced usage examples~\cite{5295283}. Nasehi et al. analyzed StackOverflow threads to understand the characteristics of good and bad code examples based on user provided votes \cite{Nasehi_what_makes}. They found that commonly down-voted API related answers lacked code examples, explanations, and shortcoming of solutions. Treude et al. proposed a machine learning based approach to find relevant API documentation from StackOverflow using both the text and metadata found from the questions and answers~\cite{Treude:2016:AAD:2884781.2884800}. Kim et al. presented a technique to automatically augment code examples from the web using code search tools to Java APIs \cite{Kim:2009:AEJ:1747491.1747552}. Jiau et al. observed a severe inequality within the context of crowd-sourced API usage examples where most of the content were related to the popular APIs \cite{jiau}. Existing research on crowd-sourced API documentation has mostly focused finding usage examples of local APIs. In this paper, we focused on REST APIs, where the API client development programming language is agnostic to API implementation language. This distinction with local APIs makes it hard to leverage the techniques proposed in the aforementioned papers for finding crowd-sourced REST API usage examples.

\subsubsection{Measuring API Usability}
Rama et al. presented a set of formulas for computing a measure of API usability based on the API's structural components such as classes, methods, parameters, return values, thread-safety~\cite{rama}. Scheller et al. presented a framework for automatically measuring the complexity of an API~\cite{scheller2015automated}. They identified a list of measurable API properties and provided formulas to compute complexities related to the interfaces, implementation and setup of an API. Grill et al. suggested using a combination of expert opinion and developer workshops to identify and collect feedback about API and its documentation related problems \cite{Grill2012}. Farooq et al. recommended using peer reviews of API code in addition to API usability studies to uncover API related bugs and incorporate feedback~\cite{Farooq:2010:AUP:1753326.1753677}. In our work, instead of measuring API usability, we have focused specifically on the relationship of REST API usability with respect to usage examples.

\subsubsection{Controlled Studies}
Nasehi et al. performed a controlled study to understand if API unit tests can be used to provide as usage examples to facilitate API client developers \cite{5609553}. The researchers found that the examples from the unit tests helped understanding the API concepts better but it was challenging for the participants to locate relevant examples. Endrikat et al. performed a controlled study with four groups of participants to understand the impact of API documentation on APIs that are implemented using programming languages with static and dynamic type systems \cite{Endrikat:2014:ADS:2568225.2568299}. Participants were given a set of failing unit tests and were asked to make them pass by writing code using the studied API. They found that the participant group using a statically typed API with explicit documentation could get the tests to pass in less time than the other groups. Ellis et al. performed a controlled study with two groups of developers to understand the impact of using a constructor vs. a factory method design pattern on API client development time \cite{Ellis:2007:FPA:1248820.1248863}. They found that API client developers needed more time to complete API tasks using factory method compared to using a constructor. Our work in this paper is based on a controlled study and shares part of the setup that were used by the aforementioned controlled studies with the following differences: our goal is to understand the obstacles faced by API client developers without usage examples;  we focus on REST APIs instead of local APIs; and, the participants are all professional software engineers.
